---
title: "Pacific Swordfish Assessment"
author: "(Your name here)"
date: "3/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(999)
```

# Comprehensive stock assessment of Western Central Pacific Swordfish

1) Calculate CPUE
2) Total Catch by nation `who is fishing for swordfish the most`?
3) Identify stocks `are there more than one?`
4) Surplus Production `estimate biomass based on parameters`
5) Comparing and predicting CPUE `How can we develop an index of relative abundance?`
6) Population Growth 
7) Mapping `To visualize the stocks`

## installing sCPUEdb package

**sCPUEdb** is a package for easily connecting and querying information found in the `pelagic` database, created by the Ferretti lab. The database is hosted by [PostgreSQL](https://www.postgresql.org/).

*Mac*
Download the `sCPUEdb_1.0.tar.gz` and then open `terminal` from your applications
-- In the terminal -- 
Run: `cd ~/Downloads` to navigate to your Downloads folder
Run: `R CMD INSTALL sCPUEdb_1.0.tar.gz` to install the package


*Windows*
Download the `sCPUEdb_1.0.tar.gz`
-- In RStudios -- 
Run: `install.packages("C:/Downloads/sCPUEdb_1.0.tar.gz", repos = NULL, type="source")`
NOTE: install any dependencies such as `mapdata` and `plyr` 
Install: Java Development Kit

```{r}
# These packages are necessary for connecting to a PostgreSQL database
# Remember to remove or comment these lines out when knitting
install.packages("RH2")
install.packages("RPostgreSQL")
```

```{r}
library(sCPUEdb)
con = connectPelagic("fmuser","swordfish") # connect to the pelagic database
code = "swo" # represents swordfish
iattc = getIATTCdataCE(con, code = code, index = "n") # Inter-American Tropical Tuna Commission
wcpfc = getWCPFCdataCE(con, code = code, index = "n") # Western Central Pacific Fisheries Commission
```

if sCPUEdb cannot be installed, use the `csv` files 
```{r}
iattc = read.csv("iattc_swo.csv")
wcpfc = read.csv("wcpfc_swo.csv")
```


### Creating CPUEs from catch and hook data and cleaning data
Like last week, we can create catch per unit effort using the `catch` and `hooks` columns in our new dataframes

```{r}
# calculate CPUE and rescale by a factor of 1000
iattc$cpue = (iattc$catch / iattc$hooks)*1000
wcpfc$cpue = (wcpfc$catch / wcpfc$hooks)*1000
```

## Total Catch 

Now that we have comprehensive CPUEs from two data sources we can do a more-informed stock assessment of swordfish in the Pacific Ocean

Next, let's look at recent catch biomass (annual) by country for IATTC and WCPFC reports. 

```{r}
library(ggplot2)
fleet_ag = aggregate(data=iattc, catch~fleet, sum) # aggregate catch by fleet
top5fleets =  fleet_ag[order(-fleet_ag$catch),] # order fleet by top catches
iattc_plot = ggplot(iattc[(iattc$year>1990 & 
                      iattc$fleet==top5fleets$fleet[1:5]),], # top fleets with most catch
       aes(fill=fleet, y=catch, x=year)) + 
    geom_bar(position="stack", stat="identity") + ggtitle("IATTC Swordfish")

fleet_ag = aggregate(data=wcpfc, catch~fleet, sum) # aggregate catch by fleet
top5fleets =  fleet_ag[order(-fleet_ag$catch),] # order fleet by top catches
wcpfc_plot = ggplot(wcpfc[(wcpfc$year>1990 & 
                      wcpfc$fleet==top5fleets$fleet[1:5]),], # top fleets with most catch
       aes(fill=fleet, y=catch, x=year)) + 
    geom_bar(position="stack", stat="identity") + ggtitle("WCPFC Swordfish")

gridExtra::grid.arrange(iattc_plot, wcpfc_plot) # arrange both plots next to each other
```

Next, let's combine the dataframes to make modeling easier
We can do this with `rbind()`

```{r}
fishcommission = rbind(iattc, wcpfc)
dim(fishcommission) # returns dimensions of dataframe or matrix/array
```

### Identify Stocks
From the 2018 WCPFC assessment, there appears to be 3 defined stocks of Pacific Swordfish, delineated by the equator and the 150W longitude line. But, let's continue looking at the northern and southern stocks we defined last week.

```{r}
dat_n = fishcommission[fishcommission$lat>0,] # subset master dataframe for ABOVE the equator
fc_north = aggregate(data=dat_n, catch~year, FUN=mean) # aggregate catch by year
fc_north$hooks = aggregate(data=dat_n, hooks~year, FUN=mean)$hooks # add aggregated hook data
fc_north$cpue = fc_north$catch/fc_north$hooks*1000 # and finally calculate average CPUE

dat_s = fishcommission[fishcommission$lat<0,] # subset master dataframe for BELOW the equator
fc_south = aggregate(data=dat_s, catch~year, FUN=mean)
fc_south$hooks = aggregate(data=dat_s, hooks~year, FUN=mean)$hooks
fc_south$cpue = fc_south$catch/fc_south$hooks*1000
```

### Surplus Production

**Exercise**
Instead of just assuming a value for `K` and `Binit`, let's use some simple statistical tools to estimate an average. Normally, you would source values for carrying capacity from literature, but here we will use random number generation to create a value from a uniform distribution.

1) use `runif()` to generate 20 random `K` values between 20,000 and 40,000 metric tons
2) take the mean of these values
3) Assume initial Biomass must be between 20% and 80% of carrying capacity. 
4) use `runif()` to generate 100 random `Binit` values 
5) calculate its mean and standard deviation

```{r}

```

```{r}
pars = c(r=0.16,K=meanK,Binit=meanBinit) # initial parameters
pars[[3]]/pars[[2]] # annual depletion based on above parameters, Binit/K
scheaferM = function(param, dat){
  B = c()
  B[1] = param['Binit']
  C=dat$catch
  for (i in 1:(nrow(dat)-1)){
    
    B[i+1]=B[i] + param['r']*B[i]*(1-B[i]/param['K']) - C[i]
  }
  B
}

biomass_north = scheaferM(param=pars, dat=fc_north) # predict biomass for northern stock
biomass_south = scheaferM(param=pars, dat=fc_south) # predict biomass for southern stock
```

```{r, fig.width=2, fig.height=2.5}
# Plot time series of predicted biomass for both stocks
# indicate minimum and maximum potential biomass / year
par(mfrow=c(2,1))
plot(biomass_north~fc_north$year, type='l', ylab="Biomass (metric tonnes)", 
     xlab="Year", main="SP of Northern Pacific Swordfish",
     ylim=(range(biomass_north)))
abline(v=fc_north$year[which(biomass_north==min(biomass_north))], col='red') # adds a vertical RED line for the year in which MINIMUM biomass was predicted
abline(v=fc_north$year[which(biomass_north==max(biomass_north))], col='green') # adds a vertical GREEN line for the year in which MAXIMUM biomass was predicted

plot(biomass_south~fc_south$year, type='l', ylab="Biomass (metric tonnes)", 
     xlab="Year", main="SP of Southern Pacific Swordfish",
     ylim=(range(biomass_south)))
abline(v=fc_south$year[which(biomass_south==min(biomass_south))], col='red')
abline(v=fc_south$year[which(biomass_south==max(biomass_south))], col='green')

```

Our predictions indicate that both stocks are doing very well as of recently.

### CPUE by stock

Let's compare CPUE of the Northern and Southern Pacific 

```{r}
par(mfrow=c(1,2))
# Northern Pacific 
plot(cpue~year, data=fc_north, type='l', xlab='Year', ylab="CPUE", main="Northern Pacific")
# Southern Pacific
plot(cpue~year, data=fc_south, type='l', xlab='Year', ylab="CPUE", main="Southern Pacific")
```

Now, let's calculate an index of relative abundance by using a catchability coefficient

$$
\text{Index of relative abundance}\\
I_t = qB_t
$$

```{r, fig.width=2, fig.height=2}
library(datalowSA)
pred_cpue = function(dat, biomass, PArea){ # dat=CPUE dataframe, biomass=SP biomass, PArea = north or south stock
cpue = dat$cpue
# estimating q from the data
qs = cpue/biomass
q = mean(qs)
pred.cpue.e = biomass*q 
plot(cpue~year, data=dat, main=paste(PArea,"Pacific Swordfish CPUE"))
pred.cpue = simpspm(pars, dat, schaefer = TRUE)
lines(pred.cpue.e~dat$year, col='red')
lines(pred.cpue~dat$year, col='black')
q
}
par(mfrow=c(2,1))
pred_cpue(fc_north, biomass_north, PArea="North")
pred_cpue(fc_south, biomass_south, PArea="South")
```

Here, it is interesting that as CPUE decreases for the Northern Pacific, fishing effort increases for the Southern Pacific, perhaps an artifact of overfishing, but our biomass trends do not support this.

```{r}
# Function for calculating depletion, Surplus Production, and harvest rates based on initial stock parameters
SPstock = function(dat, param){
require(datalowSA)
ans = displayModel(param,dat,schaefer = TRUE, addrmse = TRUE)
spmphaseplot(ans, fnt=7)
}
SPstock(fc_north,param=pars)
```

The Northern Stock is harvested sustainably according to our predictions.

```{r}
SPstock(fc_south,param=pars)
```

Predicted depletion, catch, and SP production indicate that both Pacific Swordfish stocks are being harvested sustainably.

### Population Growth

How could you estimate population size?
Can you model logistic population growth using material from Lab 4? 

K = 100000
N0 = 50000
r = 0.2

```{r}

```

### International Effort

```{r, fig.height=2.5}
library(ggplot2)

p1 = ggplot(aes(y = cpue, x = fleet), data = iattc) + geom_boxplot() +
  ylim(0,200) + ggtitle("IATTC") # create boxplot of CPUE by Nation
p2 = ggplot(aes(y = cpue, x = fleet), data = wcpfc) + geom_boxplot() +
  ggtitle("WCPFC") # create boxplot of CPUE by Nation
gridExtra::grid.arrange(p1,p2)
```

### Can we look at what else was being fished/targetted at the same time
If your sCPUEdb package is installed, you can grab IATTC and WCPFC data for other fish stocks such as:
`alb` = albacore
`bet` = big-eyed tuna
`skj` = skip-jack tuna

```{r}
code = "alb" # represents swordfish
iattc_alb = getIATTCdataCE(con, code = code, index = "n") # Inter-American Tropical Tuna Commission
wcpfc_alb = getWCPFCdataCE(con, code = code, index = "n") # Western Central Pacific Fisheries Commission
```


### Mapping

Let's try to map CPUE grids of these stocks together.





