---
title: "Intro to R part 1"
author: "Kalin Davis"
date: "1/21/2023"
output:
  word_document: default
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
  html_document: default
graphics: yes
---
```{r setup, include=F}
knitr::opts_chunk$set(echo=TRUE)
set.seed(9999)
```

This is a brief intro to R and RStudio. Here are some good resources\newline
-- VT library: The R Book, Michael J. Crawley, 2007   
-- [bookdown](https://bookdown.org/josiesmith/labbook/basics-of-r.html#objects) R basics\newline
  
Use an .Rmd file within RStudio. This way, you can experiment with code in the "code blocks" provided. When you are done, use the "Knit" option to view the final Word document.
 
## R as a calculator  

```{r}
1+1
7*4
1753/18
2+5*3
(2+5)*3
```

Simple math functions
```{r}
log(100)
sqrt(100)
```

### Objects  
You can assign values to objects easily and perform calculations.
```{r}
x <- 6
2*x + 3

y <- 2
y^2
```

Create a new object HERE
```{r}

```

### Functions
Functions simplify complex calculations
```{r}
object2 = c(5, 10, 15, 20, 25, 30) # 'c()' stands for combine  
mean(object2)
sum(object2)
```

```{r}
mean(cars$dist)
sum(cars$dist)
```
Note: Sometimes na.rm = TRUE is necessary within the paranetheses to instruct R to ignore missing data.
```{r}
mean(cars$dist, na.rm = TRUE)
```

### Vectors  
You can create vectors using the combine function `c()`
A vector is a sequence of data that are of the same type (numbers, characters)
```{r}
x <- c(1,2,3,4,5,6,7) # integers or numerals
x
class(x)
y <- c("a", "b", "c", "d") # characters
y
class(y)
```

### Indexing vectors    
Within a vector, you can extract individual values based on their index, for example:
```{r results="hold"}
x[1] # first value
x[3] # third value

y[2]
y[4]

```
  
To get multiple values, you can either extract a range of values using the colon, or by combining indices with `c()`
```{r}
x
x[2:5] # 2nd, 3rd, 4th, and 5th values
x[c(2,4,7)] # 2nd, 4th, and 7th values
```
Index a value from a vector with an object
```{r}
object3 = 3
x[object3]
```

### Lists
A list is defined as a collection of multiple data types.
```{r}
listObject <- list("Virginia Tech", 1, F)
listObject
```
  
```{r}
str(listObject)
```
  
### Matrix  
Create a matrix from using `matrix()`
```{r}
matrix(1:9, nrow = 3, ncol = 3) # range, number of rows, number of columns
```

## Basic Statistics
You can run a wide range of basic statistics in R - here we examine vector x we just created:
```{r}
x
mean(x) # arithmetic mean
var(x) # variance
sd(x) #standard deviation
min(x) # minimum
max(x) # maximum
sum(x) # sum of all values
prod(x) # product of all values
median(x) # median
```

## View summary
```{r}
summary(x) # summary statistics
```

```{r}
summary(cars)
```

### Operators
Operators create conditions and are very useful for analyzing data
```{r eval=FALSE}
x <- 5
y <- 10
x < y # less than
x <= y # less than or equal to
x > y # greater than
x >= y # greater than or equal to
x == y # is equal to
x != y #is not equal to
!x # is not x
x|y # x OR y
x&y # x AND y
isTRUE(x) # test if X is TRUE
isTRUE(TRUE)
```

## Data Frames

When multiple columns or vectors of data are combined, this is called a dataframe (think of data in a spreadsheet). Dataframes can be created manually or loaded into R using a spreadsheet in .csv format.

### Writing Dataframes  
First - We will write a simple dataset with mock catch data for a hypothetical fish species for the years 2010-2014 using `data.frame()`

```{r}
catch1 <- data.frame("year" = 2010:2014, "catch" = c(5000,5200,7100,4900,5150))
catch1
```

### Indexing Dataframes
Select value from row and column
```{r}
catch1[4,1] # row, column
```
Select row
```{r}
catch1[4,]
```
Select column
```{r}
newcatch <- catch1[1]
newcatch
```
Use the `$` to select columns as well
```{r}
catch1$year # selects all values in the 'year' column 
catch1$catch # selects all values in the 'catch' column
```

### Combining and merging dataframes
Dataframes can be merged using the row bind `rbind()` function. 
```{r}
# write another dataframe with same column names
catch2 <- data.frame("year"=c(2015:2019),  "catch"=c(4100,5700,3850,5500,4900)) # new catch dataframe
allcatch <- rbind(catch1, catch2) # binds x & y dataframes together
allcatch
```
Add a new column of **hooks deployed** with `cbind()` or `$`
```{r}
# these two lines do the same thing
cbind(allcatch, hooks = c(20000, 25000, 23000, 31000, 19000, 21000, 21000, 30000, 18000, 34000))
allcatch$hooks = c(20000, 25000, 23000, 31000, 19000, 21000, 21000, 30000, 18000, 34000)
```

Perform operations on columns
```{r}
allcatch$catch_per_hook = allcatch$catch / allcatch$hooks # divide catch by hooks deployed
allcatch
```

New data for the same years is presented
```{r}
more_catch = data.frame(year=c(2012, 2014), catch=c(4000, 2000))
more_catch
```
Use the `merge()` function to join the two dataframes 
```{r}
merged_allcatch = merge(allcatch, more_catch, by="year", all.x=T) # merge more_catch to allcatch by the year column
# all.x=T means an inner join to the first dataframe specified
merged_allcatch
```

### Subsetting Dataframes
`subset()` works by selecting a specific value or values based on a given condition
```{r}
allcatch_sub <- subset(allcatch, year>2016) # select all catch data recorded after 2016
allcatch_sub
```

```{r eval=FALSE}
# Tip: run these functions one at a time instead of the whole chunk at once 
subset(allcatch, year==2012) # only 2012 data
subset(allcatch, year!=2012) # select all EXCEPT the year 2012
subset(allcatch, year>2012 & catch>4000)
subset(allcatch, year>2012 & catch>4000 & hooks>30000)
```

## Setting a working directory and reading .csv files 

### Setting a working directory

The working directory is the directory on your computer where R will retrieve and store data as well as save outputs such as plots and spreadsheets.

For this class, you should maintain a consistent file structure such that you can always code in it. 

For example, on your local desktop, you might start a folder called "FisheriesManagement", with subfolders such as "Lab1 - Intro to R part1", "Lab2 - Intro to R part2", "Lab3 - Fisheries Data Visualization" etc., so that you can easily tell R where to save and retrieve files. There are three ways to set your working directory.

```
    |-- FisheriesManagement/                   <- root directory
        |-- Lab_1/                  
            |-- Lab1.Rmd      
            |-- Lab1.R
            |-- allcatch.csv
            |-- lab1_testdata.csv
        |-- Lab_2/
``` 

1) In the menu bar, click on "Session" and then select "Set Working Directory" from the drop-down menu. Select "Choose Directory..." and navigate to the desired folder.   
2) In your pane with the following tabs: Files, Plots, Packages, Help, and Viewer (Default position is bottom right). After selecting "Files", navigate to the desired folder and click "More" (the gear icon) and then "Set as Working Directory".
3) The command to set your working directory is `setwd()`. Everyone's home directory will not be the same.

I highly recommend to use option 3 because it is faster and does not require clicking. So, create a file structure like the one outlined above.

```{r}
getwd()
```

Here is a trick to get the directory of the file you are working on. This should typically go at the top of your Rmarkdown file
```{r}
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
getwd()
```

### Reading a .csv file
Once you have set your working directory, you can read in the data using the `read.csv()` function.
```{r}
list.files()
usa_landings <- read.csv("lab1_testdata.csv")
```

### Saving dataframes
We can now save our new master dataframe as a .csv file as well, using the `write.csv()` command. Files will be saved to your working directory.

```{r eval=FALSE}
write.csv(allcatch, "allcatch.csv", row.names = FALSE) # writes the dataframe that we created earlier
list.files()
```

### Exploring the new dataframe 
This dataset describes the total catch reported to the Food and Agriculture Organization (FAO) by the USA for the years 1950-2018.
```{r}
paste("Catch average:",mean(usa_landings$catch))
paste("Catch standard deviation:",sd(usa_landings$catch))
summary(usa_landings)
```

## Scripts
Scripts allow you to save and modify your code. You can open a script under File - New File - New R Script. 

Type your code directly into the script, and run individual lines of code either by clicking "Run" at the top of the script, or using the shortcut Cmd-Enter (Mac) / Ctrl-Enter (Windows).  

You can annotate your code with information that helps you to understand each line by commenting after the symbol `#` in a code line.

```{r}
# anything written after '#' will not be run as code
1+1 # A simple calculation
```

## Using Packages
Packages offer different objects, functions, and data
Get the list of all loaded packages
```{r}
search()
```
We are going to install `tidyverse`, which is a collection of several frequently used packages including `ggplot`, `dplyr`, and `tidyr`.  
  
```{r}
# install.packages("tidyverse")
```

```{r}
# load the package
library(tidyverse)
```

With `tidyverse` we can use more functions, such as `left_join()` 
```{r}
more_usa_landings = data.frame(year=c(1952, 1956), catch=c(40000, 20000), country="USA")
merge(usa_landings, more_usa_landings, by="year")
```

## Plotting in Base R

The Base R packages allow for some basic plotting through the `plot()` function. The default plot type will be a scatter plot. To plot the catches from our `usa_landings` dataframe, we can use the following code:

```{r}
plot(data=usa_landings, catch~year)
```

We can also use the `$` operator
```{r}
plot(usa_landings$catch~usa_landings$year)
```

Now, see what else the `plot` function can do with `?plot`
```{r}
?plot
# Select Generic X-Y Plotting and scroll down to see different arguments the function accepts  
```

This works for all functions. We can see in this case that we can change the plot style; for example, we can create the same plot as a line plot
```{r}
plot(catch~year, data=usa_landings, type="l", col='red')
```

## Plotting in `ggplot`

`ggplot` is a commonly used package for plotting as it is highly customizable. Plots are created by defining `geom` objects, such as points, lines, or polygons. There are numerous ggplot guides available which discuss how different plot components can be customized.

```{r}
ggplot(usa_landings) + # ggplot loads in the data and uses '+' to add properties
  geom_line(aes(x=year, y=catch)) + # creates the line plot
  geom_point(aes(x=year, y=catch), color="red") + # adds red points for each value
  ylab("Catch (Tons)") + # y axis title
  xlab("Year") + # x axis title
  ggtitle("United States Annual Landings") # main plot title
```

`ggplot` allows you to customize axes, labels, legends, line and shape styles, and most any other aspect of the plot you may want to modify. Look online at the available guides for additional information on plotting in ggplot. 

[bookdown_ggplot](https://bookdown.org/rdpeng/RProgDA/basic-plotting-with-ggplot2.html) 

## Using packages and functions to access data
```{r}
# install.packages("rfisheries") 
# install.packages("FSAdata")
```

```{r}
library(rfisheries)
library(FSAdata)
```

The `rfisheries` package provides access to the Open Fisheries global fisheries data - including the total annual number of *landings* for each year. The `FSA` package records biological fisheries data from the Global Fisheries Stock Assessment organization. 
```{r}
annual.landings = of_landings(species="SKJ") # Global landings of skipkack tuna
plot(data=annual.landings, catch~year, type='l')
```

What does the function `of_landings()` do?
```{r eval=FALSE}
?of_landings
```

Plot by country with a country-specific code
```{r}
annual.landings.usa = of_landings(country="AUS") # Total landings from Australia
plot(data=annual.landings.usa, catch~year, type='l')
```

Use `ggplot` to plot landings of `Red Drum` and annotate the axes and title.
```{r}
RDM_landings = of_landings(species="RDM")
ggplot(RDM_landings) +
  geom_line(aes(x=year, y=catch), col="red") +
  ylab("Annual Catch (Tons)")+
  xlab("Year")+
  ggtitle("Global Red Drum landings")
```

We can plot biological data using the `FSA` package
```{r}
data(RedDrum) # loads FSA data for Red Drum biological entries
ggplot(RedDrum) +
  geom_point(aes(x=age, y=fl), col="red") + # using points instead of a line
  ylab("Fork length (mm)")+
  xlab("Age (year)")+
  ggtitle("Red Drum length at age")
```

# Assignment instructions

Choose *1 country* and *1 species* from `rfisheries`, and *2 species* from `FSAdata`. Explore the data with statistical functions and summaries, and visualize the data with plots. Keep track of your report with RMarkdown and when you are finished, Knit the file to Word (you can knit to HTML first to see if it compiles without error).

Find all countries from `rfisheries`
```{r}
of_country_codes() # `head()` prints the top of the output, and leaves out the rest

```

Albania

Find all species from `rfisheries` 
```{r}
of_species_codes()
```
River lamprey


Use `of_landings()` with country OR species arguments

Find all species from `FSA`
[FSAdata](http://derekogle.com/fishR/data/byPackage)
Load species data with
```{r}
# data([insert species name])
# example
data(Bonito)
head(Bonito)
```

## Assignment 1

```{r}
#rfisheries

head(of_country_codes())
head(of_species_codes())
```

```{r}
annual.landings = of_landings(species = "CCT")
ggplot(annual.landings) + 
  geom_line(aes(x=year, y=catch)) + 
  geom_point(aes(x=year, y=catch), color="blue") + 
  ylab("Annual Catch (Tons)") + 
  xlab("Year") +
  ggtitle("Global Sand Tiger Shark landings")


mean(annual.landings$catch)
sd(annual.landings$catch)
summary(annual.landings$catch)

```


```{r}
annual.landings_1 = of_landings(country="FJI")
ggplot(annual.landings_1) + 
  geom_line(aes(x=year, y=catch)) + 
  geom_point(aes(x=year, y=catch), color="blue") + 
  ylab("Annual Catch (Tons)") + 
  xlab("Year") +
  ggtitle("Global Fiji landings")


mean(annual.landings_1$catch)
sd(annual.landings_1$catch)
summary(annual.landings_1$catch)


```


#FSA

```{r}
data("SLampreyGL") 
ggplot(SLampreyGL) +
  geom_point(aes(x=stock, y=recruits), col="red") + 
  ylab("Recruit")+
  xlab("Stock")+
  ggtitle("Stock and Recruitment for Sea Lamprey in Great Lakes")


x1 <- data(-1:1,20, replace=TRUE)  #had to do this to remove negative numbers
x2 <- data(-5:1,20, replace=TRUE)
df1 <- data.frame(x1,x2)
df1

summary(SLampreyGL)
sum(SLampreyGL$stock)
mean(SLampreyGL$recruits)
sd(SLampreyGL$stock)

```

```{r}
data("StripedBass1") 
ggplot(StripedBass1) +
  geom_point(aes(x=ageO, y=ageS), col="red") + 
  ylab("AgeS")+
  xlab("AgeO")+
  ggtitle("Age of Striped Bsss using Scales and Otoliths")

summary(StripedBass1)
sum(StripedBass1$ageO)
mean(StripedBass1$ageO)
mean(StripedBass1$ageS)
sd(StripedBass1$ageO)

```

