---
title: "Assignment 2"
author: "(Your name here)"
date: "1/26/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
getwd()
set.seed(9999)
library(tidyverse)
```

# Assignment 2

[DATA](https://www.fisheries.noaa.gov/alaska/commercial-fishing/alaska-groundfish-bottom-trawl-survey-data)

```{r}
nbs = read.csv("nbs1982_2019.csv") # Northern Bering Sea survey
head(nbs)
```

## Q1
Plot a histogram of bottom trawling depth and bottom temperature (remember to remove extraneous values)

```{r}
par(mfrow=c(1,2), bg = "grey54")
hist(nbs$BOT_DEPTH, xlab="Depth (m)", ylab="Hauls", main="Northern Bering Sea Trawling Survey Bottom Depth (1982-2019)", col="black", border="green", breaks=20, xlim=c(0, 80),
     ylim=c(0, 5000), cex.main=1.2)
nbs_rmtmp = nbs[!nbs$BOT_TEMP==-9999,]
hist(nbs_rmtmp$BOT_TEMP, xlab="Temperature (C)", ylab="Hauls", main="Northern Bering Sea Trawling Survey Bottom Temperature (1982-2019)", col="black", border="green", xlim=c(-5, 20))
```

## Q2
Plot the average `WTCPUE` per `HAUL` for all years.

```{r}
nbs_wtcpue = aggregate(data = nbs, WTCPUE~YEAR, FUN=mean)
head(nbs_wtcpue)

ggplot() + geom_point(data=nbs_wtcpue, aes(x=YEAR, y=WTCPUE)) + xlab("YEAR") +
  ylab("Biomass CPUE (tons)") + ggtitle("Northern Bering Sea Trawl Survey average biomass CPUE (1982-2019)") + theme_bw()


```

## Q3
Report the average and total `WTCPUE` for all vertebrates for the years `2000-2019`. *NOTE* vertebrates are identified by `SID<40000`. 

```{r}
vert_nbs = subset(nbs, SID<40000 & between(YEAR, 2000, 2019))
length(unique(vert_nbs$COMMON))
```

## Q4
Report the H-index and species evenness of all vertebrates

```{r}
fish_sum = as.data.frame(table(vert_nbs$SCIENTIFIC))
fish_sum = fish_sum[-c(1),] # From inspecting the new dataframe, the first row is empty and therefore needs to be removed
head(fish_sum)
```
```{r}
H_i = abdiv::shannon(fish_sum$Freq)
H_i

#recap
nbs = read.csv("nbs1982_2019.csv") # Northern Bering Sea survey
vert_nbs = subset(nbs, SID<40000 & between(YEAR, 2000, 2019))
fish_sum = as.data.frame(table(vert_nbs$SCIENTIFIC))
fish_sum = fish_sum[-c(1),] # From inspecting the new dataframe, the first row is empty and therefore needs to be removed
H_i = abdiv::shannon(fish_sum$Freq)
H_i

evenness = H_i/log(nrow(fish_sum))
evenness
```

## Q5 
Plot a histogram of species occurrence 

```{r}
hist(fish_sum$Freq,
  main = paste0("Species Occurences (n species=",nrow(fish_sum),")"), # Use paste and paste0 to write text and code output together
  xlab = "Number of occurrences",
  ylab = "Number of species",
  col = "bisque"
)
```

## Q6
Compare the Northern and Eastern Trawl surveys (1982-2019) by mapping both WTCPUE records for non-invertebrates. Which survey reports a relatively higher abundance?  
```{r}
ebs = read.csv("ebs1982_2019.csv") # Eastern Bering Sea survey
head(ebs)
```

```{r}

```













