---
title: "Kennedy Lakes Assessment"
author: "(Your name here)"
date: "3/23/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
set.seed(9999)
```

# Kennedy Tree Farm dataset

```{r}
dat = read.csv("kennedylakes.csv")
```

## Assignment

1) Create age classes based on length of `largemouth bass` using the Von Bertalanffy Growth Function and `rfishbase` for life history parameters.  

### rfishbase

```{r}
require(rfishbase)
LH1 = popgrowth("Micropterus salmoides") # parameters: Loo, growth coefficient k
LH2 = length_weight("Micropterus salmoides") # parameters: a, b 
```

2) Calculate and plot `catch-at-age`. 

3) Calculate total mortality rate based on the catch curve.

4) Optimize and fit a model that predicts weight of `largemouth bass` given length 
  - plot a linear fit
  - plot a non-linear fit

$$
log_eW = log_ea + b*log_eL 
$$

$$
W = aL^b
$$












