---
title: "Assignment 5"
author: "(Your name here)"
date: "2/16/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
getwd()
set.seed(9999)
require(tidyverse)
library(datalowSA)
library(FSA)
library(FSAdata)
```

1) Use negLL to estimate parameters of weight-length data for Kootenay Lake Bull Trout data from 2013. Retrieved from [github_ypr](https://poissonconsulting.shinyapps.io/shinyypr/)
  - predict weight using your optimized parameters
  - plot log weight vs. log length and fit your model
```{r}
laketrout = read.csv("laketrout.csv")
```

$$
W = aL^b
$$

Yellowfin tuna fishery data from 1957 can be found from the book: *Using R for Modelling and Quantitative Methods in Fisheries* catch, effort, and CPUE from 1934-1955
```{r}
yellow = read.csv("yellowfin1957.csv")
```

2) Plot the autocorrelation of this dataset?

3) Use the `Schaefer` and `Fox` models to predict CPUE (incorporate catch)




