---
title: "Assignment 4"
author: "(Your name here)"
date: "2/9/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
getwd()
set.seed(9999)
require(tidyverse)
require(rfishbase)
require(FSA)
require(FSAdata)
```

1) Growth Modeling
- plot an exponential growth model for a stock with `initial population size of 1500`, an `annual death rate of 0.07`, and an `annual harvest (catch) of 100` individuals. Compare the effects of different `birth rates b=0.15, b=0.2, and b=0.25` on the population trajectory. Project the population growth over `30 years`.  

```{r}

```

2) Calculate the maximum growth rate of `North Sea Haddock` (`Melanogrammus aeglefinus`) using the functions below. **HINT** inspect your data for the correct parameters!
*Parameters that we need*
- Lifespan
- Fecundity
- Age at Maturation
```{r}
# ---- Mortality rate ----------------------------

# get mortality from Then.etal.2014
getMtmax = function(lifespan){		
		M = 4.899 * lifespan^-0.916
		M
}

# get mortality using Dulvy.etal.2004's method
getMDulvy.etal.2004 = function(lifespan, agemat){
	w = (lifespan+agemat)/2 # average lifespan
	M = 1/w
	M
}

getRmax= function(method = "Smith.etal.1998", mortality = "Then.etal.2014", fec = NULL, litter = NULL, cycle = NULL, lifespan, agemat){

  # Calculate natural mortality rate
	if (mortality== "Dulvy.etal.2004") M = getMDulvy.etal.2004(lifespan = lifespan, agemat = agemat)
	if (mortality== "Then.etal.2014") M = getMtmax(lifespan = lifespan)
	if (is.numeric(mortality)) M = mortality

	if(method=="Smith.etal.1998"){
		tmax = lifespan
		tmat = agemat
		if (is.null(fec)) fec = litter/2/cycle else fec = fec
		Z<-1.5*M
	
		#this is equation 7 in the manuscript
		l_alpha<-(1-exp(-(Z)))/((fec)*(1-exp(-(Z*(tmax-tmat+1)))))

		#this is equation 6 in the manuscript
		eq6 <- function(reb) 1-exp(-(M+reb))-l_alpha*(fec)*1.25*exp(-reb*tmat)*(1-exp(-(M+reb)*(tmax-tmat+1)))
		r_Z<-uniroot(eq6, c(0,5), tol=1e-8)
		return(r_Z$root*2)
	} 
	
}
```

```{r}
hadLife = subset(popgrowth(species_list = "Melanogrammus aeglefinus"), Locality=="North Sea")
hadMat = subset(maturity(species_list = "Melanogrammus aeglefinus"), Locality=="North Sea")
hadFec = subset(fecundity(species_list = "Melanogrammus aeglefinus"), Locality=="North Sea")
```

3) Download catch and effort data for a species from your choice of an RFMO (ICCAT, IATTC, WCPFC, IOTC)
  [WCPFC](https://www.wcpfc.int/public-domain)
  [ICCAT](https://iccat.int/en/accesingdb.html)
  [IOTC](https://iotc.org/data/datasets)
  [IATTC](https://www.iattc.org/en-US/Data/Public-domain)
  - standardize CPUE (effort type and unit needs to be the same for all catches)
  - plot `CPUE` vs `year`
  
4) Calculate Z, S, and A for Florida Largemouth Bass (sampled from multiple lakes)
  - plot catch curve line

```{r}
lmbass = BassFL
```
 
