---
title: "Exercise4"
author: "(Your name here)"
date: "2/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Exercise  
Now, open a new R script, and using the `bluecatfish.csv` file on Canvas, fit a VBGF model and generate parameter estimates, and plot the data with confidence intervals generated from bootstrapping the data.

If you complete this, install the package `FSAdata`, and try these analyses on additional datasets. `FSAdata` contains length at age data for the following species (if the package is loaded, you can access the data easily, see example).
```{r eval = FALSE}
library(FSAdata)
data <- BlueCatfish
```
Other species, identified by named data frame:  
- Bonito (Australian bonito)  
- BullTroutRML2 (Bull Trout from Rocky Mountain Lakes)  
- CreekChub  
- Cabezon (Oregon)  
- Croaker2 (Atlantic Croaker)  
- DarterOnt (Channel Darters)  
- EuroPerchTH (European Perch)  
- FWDrumLE1 (Lake Erie Freshwater Drum)  
- LJCisco (Longjaw Cisco)  
- RedDrum (Atlantic Red Drum)  
- RockBassLO1 (Lake Ontario Rock Bass)  
- SardineChile  
- SpotVA2  
- StripedBass2  
- TroutBR (Brown and Rainbow Trout)  