---
title: "Population dynamics"
author: "(Your name here)"
date: "2/9/23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
getwd()
set.seed(9999)
```

You will need the following packages installed and loaded. 
```{r message=FALSE}
library(nlstools)
library(tidyverse)
library(magrittr)
library(FSA)
library(rfishbase)
```  

# Population Dynamics

## Exponential Growth model
$$
N_t = N_0e^{(b-d)t}
$$
`Nt` = population size at time
`N0` beginning population size
`b` = birth rate
`d` = death rate
`t` = time

Exponential growth of a hypothetical population using different birth and death rates.
```{r examplebox2.1}
b = c(0.2,0.2,0.2) # birth rates
d = c(0.04,0.1,0.2) # death rates
N0 = c(500,500,500) # starting populations

Time = seq(0,20,0.5) # time steps
Nt = matrix(NA, nrow = length(Time), ncol = 3)
for (i in 1:3){
  Nt[,i] = N0[i]*exp((b[i]-d[i])*Time)
}

Nt_dat = data.frame(Nt)
colnames(Nt_dat) = c('T1', 'T2', 'T3')
ggplot() + geom_line(aes(x=Time, y=Nt_dat$T1)) +
  geom_line(aes(x=Time, y=Nt_dat$T2),col=2) +
  geom_line(aes(x=Time, y=Nt_dat$T3), col=3) + 
  labs(x="Time", y="Population size (Nt)")
```

## Logistic Population Growth


$$
N_{t+1} = N_t + rN_t(1-N_t/K) - C_t 
$$
`Nt+1` = population size at timestep t+1
`Nt` = population size at time
`r` = rate of increase
`K` = carrying capacity
`C_t` = harvest rate  

```{r}
colors = c("0"="grey76", "10"="grey60", "20"="black")
r = 0.5
K = 1000
N0 = 50
Time = seq(0,25,1)
Nt = c()
Nt[1] = N0
#-----------------------
# Catch = 0
#-----------------------
for (i in 1:25){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K)
}
Nt0 = Nt
Ct = 10
#-----------------------
# Catch = 10
#-----------------------
for (i in 1:25){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
Nt1 = Nt
Ct = 20
#-----------------------
# Catch = 20
#-----------------------
for (i in 1:25){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
Nt2 = Nt

ggplot() + 
  geom_line(aes(x=Time, y=Nt0, color="0"), size=1.5) +
  geom_line(aes(x=Time, y=Nt1, color="10"), size=1.5) +
  geom_line(aes(x=Time, y=Nt2, color="20"), size=1.5) + 
  labs(x="Time", 
       y="Population size",
       color="Catch") +
    scale_color_manual(values = colors)
```
  
Manipulate the parameters k, r, and N0 to see their influence
  
### Von Bertalanffy Individual Growth

First, we will load some data in the .Dat format - you need to change the line below to direct R to the file location on your machine. Then, we will remove all negatives or missing values for length
```{r}
dat <- read.table("codAL.dat",head=TRUE)
index1<-which(dat$length>0)
dat <-dat[index1,]
```

$$
L_t = L_{\infty}(1-e^{-K(t-t_0)})
$$

#### Parameter starting values  
The Von Bertalanffy growth equation includes the parameters Linf, K, and t0. We will define estimates of these parameters to be used in a function.  As a general rule for marine fish, estimates of K=0.3 and t0=0 are a good starting point, and we can supply the maximum length in our dataset as an estimate of mean maximum length.

```{r}
parameters <- list(Linf=max(dat$length), K=0.3, t0=0)
```
#### Writing a function  
Next, we will write a function for the Von Bertalanffy growth equation.
```{r}
VBgrowth <- function(age, Linf, K, t0) Linf*(1-exp(-K*(age-t0)))
```


**Question** With our function, how can we estimate length at age 3?

```{r}

```
The package `FSA` contains a function vbFuns() which we can use to slightly modify our growth function - this allows us to supply all of the parameters as one argument.
```{r}
VBgrowth=vbFuns()
VBgrowth(3, Linf=max(dat$length), K=0.3, t0=0)
VBgrowth(3, Linf=c(max(dat$length), 0.3, 0))
```
Both of the above code lines produce the same result, but in the second, the three parameters are provided in one argument.  
  
#### Parameter estimates  
To generate estimates of our parameters, we will fit the model using the non-linear least squares method. We need to supply our function, our dataset, and our starting estimates for parameter values (which we defined above in the list `parameters`.)
```{r}
VBfit <- nls(length~VBgrowth(age, Linf, K, t0), data=dat,start=parameters)
```
To extract our parameter estimates from the list `VBfit`, we will use the `coef()` function; similarly, to get confidence intervals for each, we can use the `confint()` function.
```{r}
coef(VBfit)
confint(VBfit)
```

#### Bootstrapping  
We now have estimates based on our data, but to make our estimates more robust, we can use a technique called bootstrapping. Bootstrapping is a technique whereby the data is sampled with replacements repeatedly, and parameters are estimated from each bootstrap sample. For non-linear models (such as VBGF), we can use the function `nlsBoot()` from the package `nlstools`.
```{r}
bootFit <- nlsBoot(VBfit)
head(bootFit$coefboot)
```
We can also estimate confidence intervals from our bootstrap samples and plot them.
```{r}
confint(bootFit,plot=TRUE)
```

#### Predictions  
We can now use our updated parameter estimates to make predictions for length-at-age.
```{r}
predicts = data.frame(age=c(1:25)) #ages for predictions
predicts$length=predict(VBfit,predicts) #generate predicted length at age
```
The above predictions are based on our initial (non-bootstrapped) parameter estimates. We can also generate predictions based on the bootstrap samples (note that the `vbFuns()` function we used earlier assigns "t" as the variable for age). We will use the bootstrap data to project mean length at age 5, using the `apply()` function. `apply()` allows us to apply functions to vectors or lists; here we use the argument `MARGIN=1` to apply the function across rows, rather than columns, because each row contains a set of parameter estimates. We then use the `quantile()` function to generate a 95% confidence interval.
```{r}
predicts.age5 = apply(bootFit$coefboot, MARGIN=1, FUN=VBgrowth, t=5)
quantile(predicts.age5, c(0.025,0.975))
```

#### Plotting  
Let's plot our data.
```{r}
x = seq(0,25,length.out=500) #ages for predictions
predLength = VBgrowth(x,Linf=coef(VBfit)) #predicted length at age
xlmts = range(c(x,dat$age)) #Because we are plotting outside of our age data
ylmts = range(c(predLength,dat$length)) 
plot(length~age, data=dat, xlab="Age", ylab="Total length (cm)",
     xlim=xlmts, ylim=ylmts,pch=19,col=rgb(0,0,0,1/3))
lines(predLength~x,lwd=2)


dat2=data.frame(x=x, length=predLength)
ggplot() + geom_point(data=dat,aes(x=age, y=length))+
  geom_line(data=dat2, aes(x=x, y=predLength))+
  labs(x='Age', y='Total length (cm)')
```
  
Now we can add some confidence intervals - first we need to define the dimensions.
```{r}
LCI = UCI = numeric(length(x))
```
Now we are going to use a `for` loop to predict confidence intervals, again using the `apply()` function.
```{r}
for (i in 1:length(x)) {
  tmp = apply(bootFit$coefboot,MARGIN = 1,FUN=VBgrowth, t=x[i])
  LCI[i] = quantile(tmp,0.025)
  UCI[i] = quantile(tmp,0.975)
}
```
We'll now update our y-axis limits to include the range of the confidence intervals, and then plot again.
```{r}
ylmts <- range(c(predLength,LCI,UCI,dat$length))
plot(length~age, data=dat, xlab="Age", ylab="Total length (cm)",
     xlim=xlmts, ylim=ylmts,pch=19,col=rgb(0,0,0,1/3))
lines(predLength~x,lwd=2)
lines(UCI~x,lwd=2,lty=2, col="red")
lines(LCI~x,lwd=2,lty=2, col="red")

ggplot() + geom_point(data=dat,aes(x=age, y=length))+
  geom_line(data=dat2, aes(x=x, y=predLength))+
  labs(x='Age', y='Total length (cm)')+
  geom_line(aes(x,UCI), col='red', lty=2)+
  geom_line(aes(x,LCI), col='red', lty=2)
```

#### Check model fit  
We will check our model fit using the `residPlot()` function.
```{r}
residPlot(VBfit)
```
  
#### Log-Transform  
As you can see, as age increases, the variability of length increases (multiplicative error). If we take the natural log of each side, then we can transform the equation, fit it again using least-squares, and check the new fit.
```{r}
dat %<>% mutate(log_length=log(length))
VBLogFit = nls(log_length~log(VBgrowth(age,Linf,K,t0)),data=dat,start=parameters)
residPlot(VBLogFit)
```
