---
title: "Population dynamics, Length frequencies, and CPUEs"
author: "(Your name here)"
date: "2/9/23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lab_path = dirname(rstudioapi::getSourceEditorContext()$path)
knitr::opts_knit$set(root.dir = lab_path)
getwd()
set.seed(9999)
```

You will need the following packages installed and loaded. 
```{r message=FALSE}
library(tidyverse)
library(magrittr)
library(FSA)
library(FSAdata)
library(rfishbase)
```  

# Population Dynamics

## Exponential Growth model
$$
N_t = N_0e^{(b-d)t}
$$
`Nt` = population size at time
`N0` beginning population size
`b` = birth rate
`d` = death rate
`t` = time

Exponential growth of a hypothetical population using different birth and death rates.
```{r examplebox2.1}
b = c(0.2,0.2,0.2) # birth rates
d = c(0.04,0.1,0.2) # death rates
N0 = c(500,500,500) # starting populations

Time = seq(0,20,0.5) # time steps
Nt = matrix(NA, nrow = length(Time), ncol = 3)
for (i in 1:3){
  Nt[,i] = N0[i]*exp((b[i]-d[i])*Time)
}

Nt_dat = data.frame(Nt)
colnames(Nt_dat) = c('T1', 'T2', 'T3')
ggplot() + geom_line(aes(x=Time, y=Nt_dat$T1)) +
  geom_line(aes(x=Time, y=Nt_dat$T2),col=2) +
  geom_line(aes(x=Time, y=Nt_dat$T3), col=3) + 
  labs(x="Time", y="Population size (Nt)")
```

## Logistic Population Growth

$$
N_{t+1} = N_t + rN_t(1-N_t/K) - C_t 
$$
`Nt+1` = population size at timestep t+1
`Nt` = population size at time
`r` = rate of increase
`K` = carrying capacity
`C_t` = harvest rate  

```{r}
colors = c("0"="grey76", "10"="grey60", "20"="black")
r = 0.5
K = 1000
N0 = 50
Time = seq(0,25,1)
Nt = c()
Nt[1] = N0
#-----------------------
# Catch = 0
#-----------------------
for (i in 1:25){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K)
}
Nt0 = Nt
Ct = 10
#-----------------------
# Catch = 10
#-----------------------
for (i in 1:25){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
Nt1 = Nt
Ct = 20
#-----------------------
# Catch = 20
#-----------------------
for (i in 1:25){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
Nt2 = Nt

ggplot() + 
  geom_line(aes(x=Time, y=Nt0, color="0"), size=1.5) +
  geom_line(aes(x=Time, y=Nt1, color="10"), size=1.5) +
  geom_line(aes(x=Time, y=Nt2, color="20"), size=1.5) + 
  labs(x="Time", 
       y="Population size",
       color="Catch") +
    scale_color_manual(values = colors)
```
  
Now we look at intrinsic growth rates > 1 which should lead to unstable and oscillatory population dynamics. 
```{r fig2.6}
r = 0.25
K = 1000
N0 = 1
Time = seq(0,100,1)
Nt = c()
Nt[1] = N0
Ct = 0
for (i in 1:100){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
par(mfrow  = c(2,2))
plot(Nt~Time, type = "l", ylim = range(Nt), main="r=0.25")
r = 1.85
for (i in 1:100){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
plot(Nt~Time, type = "l", ylim = range(Nt), main="r=1.85")
r = 2.5
for (i in 1:100){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
plot(Nt~Time, type = "l", ylim = range(Nt), main="r=2.5")
r = 3
for (i in 1:100){
  Nt[i+1] = Nt[i] + r*Nt[i]*(1-Nt[i]/K) - Ct
}
plot(Nt~Time, type = "l", ylim = range(Nt), main="r=3")
```

### Predicting growth rate of a population
*Parameters that we need*
- Lifespan
- Fecundity
- Age at Maturation
```{r}
# ---- Mortality rate ----------------------------

# get mortality from Then.etal.2014
getMtmax = function(lifespan){		
		M = 4.899 * lifespan^-0.916
		M
}

# get mortality using Dulvy.etal.2004's method
getMDulvy.etal.2004 = function(lifespan, agemat){
	w = (lifespan+agemat)/2 # average lifespan
	M = 1/w
	M
}

getRmax= function(method = "Smith.etal.1998", mortality = "Then.etal.2014", fec = NULL, litter = NULL, cycle = NULL, lifespan, agemat){

  # Calculate natural mortality rate
	if (mortality== "Dulvy.etal.2004") M = getMDulvy.etal.2004(lifespan = lifespan, agemat = agemat)
	if (mortality== "Then.etal.2014") M = getMtmax(lifespan = lifespan)
	if (is.numeric(mortality)) M = mortality

	if(method=="Smith.etal.1998"){
		tmax = lifespan
		tmat = agemat
		if (is.null(fec)) fec = litter/2/cycle else fec = fec
		Z<-1.5*M
	
		#this is equation 7 in the manuscript
		l_alpha<-(1-exp(-(Z)))/((fec)*(1-exp(-(Z*(tmax-tmat+1)))))

		#this is equation 6 in the manuscript
		eq6 <- function(reb) 1-exp(-(M+reb))-l_alpha*(fec)*1.25*exp(-reb*tmat)*(1-exp(-(M+reb)*(tmax-tmat+1)))
		r_Z<-uniroot(eq6, c(0,5), tol=1e-8)
		return(r_Z$root*2)
	} 
	
}
```

Look at Spiny Dogfish in the Northwest Atlantic.
```{r}
spinyLife = popgrowth(species_list = c("Squalus acanthias"))
```

```{r}
spinyLife = subset(popgrowth(species_list = c("Squalus acanthias")), Locality=="Northwestern Atlantic")
spinyFec = fecundity(species_list = c("Squalus acanthias"))[1,]
spinyMat = subset(maturity(species_list = c("Squalus acanthias")), Locality=="Northwest Atlantic")
```

Calculate growth rate for Northwest Atlantic spiny dogfish stock
```{r}
lifespan = mean(spinyLife$tmax, na.rm=T)
fecundity = (spinyFec$FecundityMin+spinyFec$FecundityMax) / 2 # mean of minimum and maximum fecundity
agemat = (spinyMat$AgeMatMin + spinyMat$AgeMatMin2) / 2 # mean of two records for minimum age at maturity
rmax = getRmax(method = "Smith.etal.1998", mortality = "Then.etal.2014", fec = fecundity, lifespan = lifespan, agemat = agemat)
paste0("Intrinsic Growth Rate = ", round(rmax, 3))
```

## Length frequency distributions

Growth increment data for Red Lakes Walleye (Sander vitreus) 
```{r}
data = WalleyeRL
hist(~tl,data=bg11,breaks=seq(20,270,10),
xlab="Total Length (mm)")
```

## Calculating CPUEs
Catch and effort of swordfish by longline in the Indian Ocean and the Western Central Pacific 
```{r}
iattc = read.csv("iattc_swo.csv") # IATTC catch and effort for swordfish
wcpfc = read.csv("wcpfc_swo.csv") # WCPFC catch and effort for swordfish
```

### Creating CPUEs from catch and hook data 
Create catch per unit effort using the `catch` and `hooks` columns 

```{r}
# calculate CPUE and rescale by a factor of 1000
iattc$cpue = (iattc$catch / iattc$hooks)*1000
wcpfc$cpue = (wcpfc$catch / wcpfc$hooks)*1000
```

## Total Catch by Nation

Now that we have comprehensive CPUEs from two data sources we can do a more-informed stock assessment of swordfish in the Pacific Ocean

Next, let's look at recent catch biomass (annual) by country for IATTC and WCPFC reports. 

```{r}
# IATTC
fleet_ia = aggregate(data=iattc, catch~fleet, sum) # aggregate catch by fleet
top5fleets =  fleet_ia[order(-fleet_ia$catch),] # order fleet by top catches
iattc_plot = ggplot(iattc[(iattc$year>1990 & iattc$fleet==top5fleets$fleet[1:5]),], aes(fill=fleet, y=catch*0.001, x=year)) + # top fleets with most catch
    geom_bar(position="stack", stat="identity") + ggtitle("IATTC Swordfish") + 
  ylab("Catch (t)")

# WCPFC
fleet_wc = aggregate(data=wcpfc, catch~fleet, sum) # aggregate catch by fleet
top5fleets =  fleet_wc[order(-fleet_wc$catch),] # order fleet by top catches
wcpfc_plot = ggplot(wcpfc[(wcpfc$year>1990 & wcpfc$fleet==top5fleets$fleet[1:5]),], aes(fill=fleet, y=catch*0.001, x=year)) + # top fleets with most catch
    geom_bar(position="stack", stat="identity") + ggtitle("WCPFC Swordfish") +
  ylab("Catch (t)")
```

```{r}
gridExtra::grid.arrange(iattc_plot, wcpfc_plot) # arrange both plots next to each other
```

### Identify Stocks
From the 2018 WCPFC assessment, there appears to be 3 defined stocks of Pacific Swordfish, delineated by the equator and the 150W longitude line. 

Look at the `North` and `South` swordfish stocks in the Western Pacific
```{r}
dat_n = wcpfc[wcpfc$lat>0,] # subset master dataframe for ABOVE the equator
fc_north = aggregate(data=dat_n, catch~year, FUN=mean) # aggregate catch by year
fc_north$hooks = aggregate(data=dat_n, hooks~year, FUN=mean)$hooks # add aggregated hook data
fc_north$cpue = fc_north$catch/fc_north$hooks*1000 # and finally calculate average CPUE

dat_s = wcpfc[wcpfc$lat<0,] # subset master dataframe for BELOW the equator
fc_south = aggregate(data=dat_s, catch~year, FUN=mean)
fc_south$hooks = aggregate(data=dat_s, hooks~year, FUN=mean)$hooks
fc_south$cpue = fc_south$catch/fc_south$hooks*1000
```


```{r}
north = ggplot() + 
  geom_line(data=fc_north, aes(x=year, y=cpue)) +
  ylab("CPUE (kg/hook)") +
  ggtitle("Northwest Pacific Swordfish") +
  theme(panel.grid = element_line(colour = "black",size = 0.05),
        panel.border = element_rect(fill = NA))

south = ggplot() + 
  geom_line(data=fc_south, aes(x=year, y=cpue)) +
  ylab("CPUE (kg/hook)") +
  ggtitle("Southwest Pacific Swordfish") +
  theme(panel.grid = element_line(colour = "black",size = 0.05),
        panel.border = element_rect(fill = NA))

gridExtra::grid.arrange(north, south)
```


